**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2020-09-25 manuscript/ecosphere_resubmit/manuscript.Rmd
last resubmission commit

---
title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze for a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus does show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

>>>>

---
title: "**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze for a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus does show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
